Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-02 sections/abstract.tex
updates

\begin{abstract}
\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.  Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.  However, arrays can only handle dense rectilinear integer grids. Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry. We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression. }

\HIDE{
Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler technology.
Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes. So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

In this paper, we describe a compiler, Finch,
which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
sparse tensors, databases, and lossless compression.
We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.}

From FORTRAN to Numpy, arrays have revolutionized how we express computation. However, arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry.
Support for structured data is fragmented and incomplete.
Existing frameworks limit the array structures and program control flow they support to better simplify the problem.

>>>>

\begin{abstract}
From FORTRAN to Numpy, arrays have revolutionized how we express computation. However, arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry.
Support for structured data is fragmented and incomplete.
Existing frameworks limit the array structures and program control flow they support to better simplify the problem.

### 2024-04-02 sections/background.tex
Update on Overleaf.
updates

\section{Background On Looplets}
Finch represents iteration patterns using Looplets, a language that decomposes
datastructure iterators hierarchically. 
%
Looplets represent the control-flow
structures needed to iterate over any given datastructure, or multiple
datastructures simultaneously. 
%
In particular, looplets are good at lifting code
to the highest loop level that it's needed and subdividing iteration
hierarchically in coordinate space.
%
Because looplets are compiled with
progressive lowering, structure-specific mathematical optimizations such as
integrals, multiply by zero, etc. can be implemented using simple compiler
passes like term rewriting and constant propagation during the intermediate
lowering stages. \cite{ahrens_looplets_2023}

In this work, we partially simplify the presentation of Looplets to focus on the
semantics, rather than precise implementation. Since Finch is designed to be
embedded into another language\saman{not clear}, we allow for several of the looplets to utilize
functions that modify the state of that language. It is assumed that if a
looplet introduces a variable to be used in a child looplet, the child looplet
will not modify that variable.\saman{No context yet to understand this. Need to introduce what looplets are (Fig 2)!}

@@

    \begin{minipage}[c]{0.65\linewidth}
        $\finchstepper([seek], next, body)$ The stepper looplet
        represents a variable number of looplets, concatenated. Since our
        looplets may be skipped over due to conditions or various rewrites, the
        $seek$ function allows us to fast-forward the state to the start of the
        root loop extent when it comes time to lower the stepper. The $next$
        function advances the state to the next iteration of the stepper. 
    \end{minipage}%
    \begin{minipage}[c]{0.35\linewidth}
        \centering
        \includegraphics[scale=0.20]{Looplets-stepper.png}
    \end{minipage}
    \caption{The looplet language, as understood in a correct execution of a Finch program.}
\end{figure}

>>>>

\section{Background On Looplets}
Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. 
%
Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously. 
%
In particular, looplets are good at lifting code to the highest loop level that it's needed and subdividing iteration hierarchically in coordinate space.
%
Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc. can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages \cite{ahrens_looplets_2023}.

The looplets are described in Figure~\ref{fig:looplets}. We simplify the presentation to focus on the semantics, rather than precise implementation.  Several looplets introduce or modify variables in the scope of the target language. It is assumed that if a looplet introduces a variable to be used in a child looplet, the child looplet will not modify that variable.

@@

    \begin{minipage}[c]{0.65\linewidth}
        $\finchstepper([seek], next, body)$ The stepper looplet
        represents a variable number of looplets, concatenated. Since our
        looplets may be skipped over due to conditions or various rewrites, the
        $seek$ function allows us to fast-forward the state to the start of the
        root loop extent when it comes time to lower the stepper. The $next$
        function advances the state to the next iteration of the stepper. 
    \end{minipage}%
    \begin{minipage}[c]{0.35\linewidth}
        \centering
        \includegraphics[scale=0.20]{Looplets-stepper.png}
    \end{minipage}
    \caption{The looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
\end{figure}

### 2024-04-02 sections/cut.tex
updates

% \teo{
% Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
% %
% Finch supports \textit{both} flexible programming constructs and diverse data structures. 
% %
% Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
% %
% The Finch compiler uses the structure
% of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}

>>>>

\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.  Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.  However, arrays can only handle dense rectilinear integer grids. Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry. We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression. }

\HIDE{
Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler technology.
Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes. So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

In this paper, we describe a compiler, Finch,
which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
sparse tensors, databases, and lossless compression.
We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.}
% \teo{
% Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
% %
% Finch supports \textit{both} flexible programming constructs and diverse data structures. 
% %
% Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
% %
% The Finch compiler uses the structure
% of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}

### 2024-04-02 sections/intro.tex
Update on Overleaf.
updates

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.
% EASY CUT
%These frameworks have been the
%subject of extensive industry investment to enable performant implementations,
%and often operate at the peak capacity of the hardware they run on
%\cite{lo_roofline_2015}.

The success and ubiquity of arrays is largely due to their simplicity. 
%
Since their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
%
Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
%
%This
%simplicity enables extensive interoperability, convenience layers, and
%optimizations by breaking the abstraction barrier between array representation
%and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectorization,
loop-carry-dependencies, and more. Several optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.
%
However, while dense arrays are the easiest way to program for performance, the world is not all dense.

@@

%
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
%
This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
%

@@

\begin{table}[h!]
\noindent
\flushleft
\begin{minipage}[t]{.49\textwidth}
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Einsums/Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  % Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Unrestricted Read/Write              &    \checkmark        &  &            &  &            &  \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
\end{minipage} 
\begin{minipage}[t]{.49\textwidth}
  \flushright
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse Operand              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Multiple Sparse  Operands                 &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{minipage}
\end{table}

>>>>

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern array programming languages such as NumPy~\cite{harris_array_2020},
SciPy~\cite{virtanen_scipy_2020}, MatLab~\cite{moler_history_2020},
TensorFlow~\cite{abadi_tensorflow_2016}, PyTorch~\cite{paszke_pytorch_2019}, and
Halide~\cite{ragan-kelley_halide_2013} have pushed the limits of productive data
processing with arrays, fueling breakthroughs in machine learning, scientific
computing, image processing, and more.
% EASY CUT
%These frameworks have been the
%subject of extensive industry investment to enable performant implementations,
%and often operate at the peak capacity of the hardware they run on
%\cite{lo_roofline_2015}.

The success and ubiquity of arrays is largely due to their simplicity. 
%
Since their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
%
Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
%
%This
%simplicity enables extensive interoperability, convenience layers, and
%optimizations by breaking the abstraction barrier between array representation
%and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectorization,
loop-carry-dependencies, and more. A myriad of optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.
%
However, while dense arrays are the easiest way to program for performance, the world is not all dense.

@@

%
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure/data set and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{data structure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
%
However, libraries will only ever support a subset of
programs on a subset of data structure combinations.
%
Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semi-rings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
%
Special-purpose
compilers like TACO~\cite{kjolstad_tensor_2019}, Taichi~\cite{hu_taichi_2019}, StructTensor~\cite{ghorbani2023compiling}, or CoRa~\cite{fegade_cora_2022} which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions.
%
This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
%

@@

\begin{table}[h!]
\noindent
\flushleft
\begin{minipage}[t]{.49\textwidth}
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Einsums/Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
  % Automatic API Fusion     &            &            &            &            &            & \checkmark \\
%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            &           \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Unrestricted Read/Write              &    \checkmark        &  &            &  &            &  \\
  \end{tabular}
  \caption{Control flow support across various tools.}
  \label{tab:features}
\end{minipage} 
\begin{minipage}[t]{.49\textwidth}
  \flushright
  \scriptsize
  \begin{tabular}{l|cccccc}
  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            &            & \checkmark \\
  One Sparse Operand              &            & \checkmark &            & \checkmark &            &\checkmark \\
  Multiple Sparse  Operands                 &            & \checkmark &            &            &            &\checkmark \\
  Run-length               &            &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
  \end{tabular}
  \caption{Data structure support across various tools.  Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{minipage}
\vspace{-24pt}
\end{table}

### 2024-04-02 sections/tensors.tex
Update on Overleaf.
updates

\subsection{Scalars}

\subsubsection{Sparse Scalars}
\subsubsection{Early Break Scalars}

@@

    \hline
    \textbf{Sequential Reading and Updating} \\
    \hline
    \textbf{Dense}:
    The dense format is the simplest format, mapping $fiber(l, p)[i] \rightarrow fiber(l.lvl, p \times l.shape + i)$.
    %
    This format is used to store dense data and is often a convenient format for the root level of a tensor.
    %
    Due to its simplicity, freezing and thawing the level are no-ops. \\
    \textbf{DenseRLE}:
    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $qth$ run in the $p^th$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
    %
    A challenge arises for this level: it is difficult to merge duplicate runs.
    %
    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
    %
    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
    %
    Instead, we merge the duplicates during the freeze phase.
    %
    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer. \\
    %
    %It also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseList}:
    The simplest sparse format, used to construct popular formats like CSR, CSC, DCSR, DCSC, and CSF.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. \\
    \textbf{SparsePinpoint}:
    Similar to SparseList, but only one nonzero in each subfiber, eliminating the need for the $ptr$ field.
    %
    It stores a vector $idx$, such that $idx[p]$ is the nonzero index in the subfiber at position $p$. \\
    \textbf{SparseRLE}:
    Similar to DenseRLE level, but because runs are sparse, we must also store the start of each run.
    %
    It stores three vectors $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
    %
    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseInterval}:
    Similar to SparseRLE, but only stores one run per subfiber, eliminating the need for the $ptr$ field.
    %
    This level does not deduplicate as it cannot store intermediate results with more than one run.
    %
    It stores two vectors, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively. \\
    \textbf{SparseVBL}: Used to represent blocked data. It stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$. \\
    \textbf{SparseBand}: Similar to SparseVBL, but stores only one block per subfiber, eliminating the need for the $ptr$ field. It stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$. \\

    \hline
    \textbf{Random Access Reading and Updating} \\
    \hline
    \textbf{SparseHash}:
    The sparse hash format uses a hash table to store the locations of nonzeros, and sorts the unique indices for iteration during the freeze phase.
    %
    This allows for efficient random access, but not incremental construction, as the freeze phase runs in time proportional to the number of nonzeros in the entire level.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. 
    %
    Also stores a hash table $tbl$ for construction and random access in the level. \\
    \textbf{SparseBytemap}
    The SparseBytemap format uses a bytemap to store which locations have been written to.
    %
    Unlike the SparseHash format, the bytemap assembles the entire space of possible subfibers.
    %
    This accelerates random access in the format, but requires a high memory overhead.
    %
    Because we don't want to reallocate all of the memory in each iteration, the declaration of this format instead re-assembles only the dirty locations in the tensor.
    %
    This format is analogous to the default workspace format used by TACO.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
    %
    These vectors are used to collect dirty locations.
    %
    It also stores $tbl$, a dense array of booleans such that $tbl[shape * p + i]$ is true when there is a nonzero at index $i$ in the subfiber at position $p$. \\
    \hline
    \textbf{Leaf Levels} \\
    \hline
    \textbf{Element}:
    The element level uses an array $val$ to store a value for each position $p$. The zero (fill) value is configurable.\\
    \textbf{Pattern}:
    The pattern level statically represents a leaf level with a fill value of $false$ and whose stored values are all $true$. \\
    \hline
    \end{tabular}
    \caption{The main level formats supported by Finch. Note that all non-leaf
    levels store a the dimension of the subfibers and a child level. Since we
    must be able to handle the case where a sublevel is not stored because a
    parent level is sparse, all of Finch's sparse formats use a dirty bit during
    writing to determine whether the sublevel has been modified from it's
    default fill value and thus, whether it needs to be stored.}
    \label{table:formats}
\end{table}

>>>>

\subsection{Scalars}
Because elements are geared towards representing many leaves in a level, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
%
Scalars can also affect how the leaves are lowered.
%
Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
%
Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
%
We also provide early break scalars, which modify the stepper looplets to
re-specialize the loop whenever a reduction into an early break scalar hits an
annihilator.
%
We assume that we don't need to re-specialize other looplets because the stepper
is the only one which repeats a non-constant number of times.

@@

    \hline
    \textbf{Sequential Access Levels} \\
    \hline
    \textbf{Dense}:
    The dense format is the simplest format, mapping $fiber(l, p)[i] \rightarrow fiber(l.lvl, p \times l.shape + i)$.
    %
    This format is used to store dense data and is often a convenient format for the root level of a tensor.
    %
    Due to its simplicity, freezing and thawing the level are no-ops. \\
    \textbf{DenseRLE}:
    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $qth$ run in the $p^th$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
    %
    A challenge arises for this level: it is difficult to merge duplicate runs.
    %
    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
    %
    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
    %
    Instead, we merge the duplicates during the freeze phase.
    %
    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer. \\
    %
    %It also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseList}:
    The simplest sparse format, used to construct popular formats like CSR, CSC, DCSR, DCSC, and CSF.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. \\
    \textbf{SparsePinpoint}:
    Similar to SparseList, but only one nonzero in each subfiber, eliminating the need for the $ptr$ field.
    %
    It stores a vector $idx$, such that $idx[p]$ is the nonzero index in the subfiber at position $p$. \\
    \textbf{SparseRLE}:
    Similar to DenseRLE level, but because runs are sparse, we must also store the start of each run.
    %
    It stores three vectors $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
    %
    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\
    \textbf{SparseInterval}:
    Similar to SparseRLE, but only stores one run per subfiber, eliminating the need for the $ptr$ field.
    %
    This level does not deduplicate as it cannot store intermediate results with more than one run.
    %
    It stores two vectors, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively. \\
    \textbf{SparseVBL}: Used to represent blocked data. It stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$. \\
    \textbf{SparseBand}: Similar to SparseVBL, but stores only one block per subfiber, eliminating the need for the $ptr$ field. It stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$. \\

    \hline
    \textbf{Random Access Levels} \\
    \hline
    \textbf{SparseHash}:
    The sparse hash format uses a hash table to store the locations of nonzeros, and sorts the unique indices for iteration during the freeze phase.
    %
    This allows for efficient random access, but not incremental construction, as the freeze phase runs in time proportional to the number of nonzeros in the entire level.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. 
    %
    Also stores a hash table $tbl$ for construction and random access in the level. \\
    \textbf{SparseBytemap}
    The SparseBytemap format uses a bytemap to store which locations have been written to.
    %
    Unlike the SparseHash format, the bytemap assembles the entire space of possible subfibers.
    %
    This accelerates random access in the format, but requires a high memory overhead.
    %
    Because we don't want to reallocate all of the memory in each iteration, the declaration of this format instead re-assembles only the dirty locations in the tensor.
    %
    This format is analogous to the default workspace format used by TACO.
    %
    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
    %
    These vectors are used to collect dirty locations.
    %
    It also stores $tbl$, a dense array of booleans such that $tbl[shape * p + i]$ is true when there is a nonzero at index $i$ in the subfiber at position $p$. \\
    \hline
    \textbf{Leaf Levels} \\
    \hline
    \textbf{Element}:
    The element level uses an array $val$ to store a value for each position $p$. The zero (fill) value is configurable.\\
    \textbf{Pattern}:
    The pattern level statically represents a leaf level with a fill value of $false$ and whose stored values are all $true$. \\
    \hline
    \textbf{Scalars} \\
    \hline
    \textbf{Scalar}:
    A dense scalar that, unlike a variable, supports reduction.  \\
    \textbf{SparseScalar}:
    A scalar with a dirty bit which specializes on the fill value when it occurs.\\
    \textbf{EarlyBreakScalar}:
    A scalar which triggers early breaks in reductions whenever an annihilator is encountered.\\
    \hline
    \end{tabular}
    \caption{The main level formats supported by Finch. Note that all non-leaf
    levels store a the dimension of the subfibers and a child level. Since we
    must be able to handle the case where a sublevel is not stored because a
    parent level is sparse, all of Finch's sparse formats use a dirty bit during
    writing to determine whether the sublevel has been modified from it's
    default fill value and thus, whether it needs to be stored.}
    \label{table:formats}
\end{table}
